---
title: "Example of use vaccineff: static matching"
author: "David Santiago Quevedo"
date: "2024-03-12"
output:
  html_document:
    css: style.css
  pdf_document: default
image: null
editor_options:
  markdown:
    wrap: 72
licenses: "CC-BY"
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE) 
```

# Use case 

```{R}
library("vaccineff")
library("ggplot2")
library("cowplot")
data(cohortdata)
head(cohortdata)
```

## Data preparation



* End date defined for the study:
```{r}
end_cohort <- as.Date("2044-12-31")
```

* Dose selected for the analysis: `date_vaccine_2`
```{r}
vacc_date_col <- "vaccine_date_2"
```

* Immunization delay given by time to protecting effect: `14 days`
```{r}
immunization_delay <- 14
```

* Outcome selected for the analysis
```{r}
outcome <- "death_date"
```

* Construct immunization date for every register based on previous definitions
```{r}
cohortdata$immunization <-
  get_immunization_date(
    data = cohortdata,
    outcome_date_col = outcome,
    outcome_delay = 0,
    immunization_delay = immunization_delay,
    vacc_date_col = vacc_date_col,
    end_cohort = end_cohort,
    take_first = FALSE)
```

* Beginning of follow-up period is defined based on the minimum immunization 
date of the cohort  
```{r}
start_cohort <- as.Date(min(cohortdata$immunization, na.rm = TRUE))
```

* Add vaccine status
```{r}
cohortdata$vaccine_status <- set_status(
  data = cohortdata,
  col_names = "immunization",
  status = c("v", "u")
)
```

## Matching population
```{r}
matched_cohort <- match_cohort(data = cohortdata,
  status_vacc_col = "vaccine_status",
  nearest = c(age = 1)
)

head(matched_cohort[order(matched_cohort$subclass), ], n = 4)
```

The couples with one or both members having a censoring date must be completely 
censored. To do it, the minimum censoring date between the two members is
imputed to both of them. Imputation occurs only if the date to be imputed is 
lower than the outcome date of each member.
```{r}
censoring_date <- "death_other_causes"
matched_cohort$censoring_date <-  get_censoring_date_match(
  data = matched_cohort,
  outcome_date_col = "death_date",
  censoring_date_col = censoring_date
)
```

### FUNCTION TO MATCH INFORMATION BETWEEN COUPLES

Other information must be imputed to partners. The following function
was designed to impute the minimum value of a given variable to each couple

```{r}
match_info <- function(data,
                       info_to_match) {
  matched_info <- unlist(
    tapply(data[[info_to_match]],
      data$subclass,
      function(x) {
        if (all(is.na(x))) {
          return(NA)
        } else {
          return(as.character(min(x, na.rm = TRUE)))
        }
      }
    )
  )
  # return data matched by subclass
  return(matched_info[data$subclass])
}
```

### Time-to-event and beginning of follow-up period

The beginning of the follow-up period of every couple must be the
immunization date of the vaccinated member.
```{r}
matched_cohort$t0_fu <- as.Date(match_info(data = matched_cohort,
  info_to_match = "immunization"
))
```

Time-to-event is calculated starting from the beginning of the follow-up
period of each couple.
```{r}
matched_cohort$time_to_death <- get_time_to_event(
  data = matched_cohort,
  outcome_date_col = "death_date",
  censoring_date_col = "censoring_date",
  start_cohort = start_cohort,
  end_cohort = end_cohort,
  start_from_immunization = TRUE,
  immunization_date_col = "t0_fu"
)
```

The matched strategy is not rolling-calendar. This generates matches between
individuals that could present an outcome before the immunization date of their
match. A first strategy to solve this is removing the problematic matches.
To do this it is necessary to match the negative time to event for every 
couple.

```{r}
matched_cohort$matched_time <- match_info(data = matched_cohort,
  info_to_match = "time_to_death"
)
neg_times <- matched_cohort[matched_cohort$matched_time < 0, ]

head(neg_times[order(neg_times$subclass), ], n = 4)
```

The couples with negative `matched_time` in the previous dataframe are removed.
```{r}
cohort_corrected <- matched_cohort[matched_cohort$matched_time > 0, ]
print(paste0("Number of matched registers: ", nrow(matched_cohort)))
print(paste0("Number of registers with correct time-to-event: ", nrow(cohort_corrected)))
```

### Outcome Status

The event status is calculated based on the outcome date. For the registers 
with censoring date, it is forced to be 0 (negative).

```{r}
# set outcome statues
cohort_corrected$death_status <- set_status(
  data = cohort_corrected,
  col_names = outcome
)

cohort_corrected$death_status <- ifelse(
  !is.na(cohort_corrected$censoring_date) &
    !is.na(cohort_corrected$death_date) &
    (cohort_corrected$censoring_date <= cohort_corrected$death_date),
  yes = 0,
  no = cohort_corrected$death_status
)
```

## Estimations

### Survival curve
The survival curve for the matched cohort 
```{r}
cohort_corrected$vaccine_status <- as.character(cohort_corrected$vaccine_status)

plt <- plot_survival(
  data = cohort_corrected,
  outcome_status_col = "death_status",
  time_to_event_col = "time_to_death",
  vacc_status_col = "vaccine_status",
  vaccinated_status = "v",
  unvaccinated_status = "u",
  vaccinated_color = "steelblue",
  unvaccinated_color = "darkred",
  start_cohort = start_cohort,
  end_cohort = end_cohort,
  percentage = TRUE,
  cumulative = TRUE
)
plt
```

```{r}
surv <- plt$data

surv[surv$time == max(surv$time), ]
1-(0.001370489/0.002667872)
```

The proportional hazards is checked by means of the loglog curve using the 
information from the survival curve

```{r}
surv$loglog <- log(-log(surv$surv))

surv$logtime <- log(surv$time)
plt_loglog <- ggplot(data = surv) +
geom_step(ggplot2::aes(x = .data$logtime,
                       y = .data$loglog,
                       color = .data$strata)
  ) +
  theme_classic() +
  labs(x = "Log[Time to event] (Days)",
       y = "Log[-Log[Survival probability]]") +
  labs(colour = "Vaccine Status") +
  scale_color_manual(
    name = "Vaccine Status",
    values = c("steelblue", "darkred"),
    labels = c("v", "u")
  )
plt_loglog
```

### Cox Model
 
 
```{r}
p_thr <- 0.05

cx <- survival::coxph(  # nolint
  survival::Surv(time_to_death, death_status) ~
    vaccine_status , data = cohort_corrected
)

# Test the Proportional Hazards Assumption
test <- survival::cox.zph(cx)
hr <- round(exp(stats::coef(cx)), digits = 4)

# extract first and second element as limits
ci025 <- round(exp(stats::confint(cx)), 4)[1]
ci975 <- round(exp(stats::confint(cx)), 4)[2]

p <- test$table["GLOBAL", "p"]

if (p < p_thr) {
  ph <- "reject"
} else {
  ph <- "accept"
}
df_summ <- data.frame(
  HR = hr,
  HR_low = ci025,
  HR_high = ci975,
  V_eff = 1 - hr,
  V_eff_low = 1 - ci975,
  V_eff_high = 1 - ci025,
  PH = ph,
  p_value = p # p_value must be a numeric
)

print(df_summ)
```
 
 Including the prop_score in the regression blows the upper value of the CI.
 Additionally, it over-estimates the effectiveness and keeps the same the
 p-value. It is not adding any level of accuracy to the model.
 
```{r}
p_thr <- 0.05

cx <- survival::coxph(  # nolint
  survival::Surv(time_to_death, death_status) ~
    vaccine_status + prop_score, data = cohort_corrected
)

# Test the Proportional Hazards Assumption
test <- survival::cox.zph(cx)
hr <- round(exp(stats::coef(cx)), digits = 4)

# extract first and second element as limits
ci025 <- round(exp(stats::confint(cx)), 4)[1]
ci975 <- round(exp(stats::confint(cx)), 4)[2]

p <- test$table["GLOBAL", "p"]

if (p < p_thr) {
  ph <- "reject"
} else {
  ph <- "accept"
}
df_summ <- data.frame(
  HR = hr,
  HR_low = ci025,
  HR_high = ci975,
  V_eff = 1 - hr,
  V_eff_low = 1 - ci975,
  V_eff_high = 1 - ci025,
  PH = ph,
  p_value = p # p_value must be a numeric
)

print(df_summ)
```
 
 
 